Article Title

### 2014-06-21 plos.tex
updated the statistical tests for Fig. 6.

\begin{document}
% Title must be 150 characters or less
\begin{flushleft}
{\Large
Improved estimation and interpretation of correlations in neural circuits
}
% Insert Author names, affiliations and corresponding author email.
\\
Dimitri Yatsenko,$^{1}$, 
Kre\v{s}imir Josi\'{c}$^{2}$,
Alexander S.~Ecker$^{1,3,4}$,
Emmanouil Froudarakis$^{1}$,
R.~James Cotton$^{1}$,
Andreas S.~Tolias$^{1,5,\ast}$
\\
\bf{1} Department of Neuroscience, Baylor College of Medicine, Houston, TX, USA
\\
\bf{2} Department of Mathematics and Department of Biology and Biochemistry, University of Houston, Houston, TX, USA
\\
\bf{3}  Werner Reichardt Center for Integrative Neuroscience and Institute for Theoretical Physics, University of T\"ubingen, Germany
\\
\bf{4} Bernstein Center for Computational Neuroscience, T\"ubingen, Germany
\\
\bf{5} Department of Computational and Applied Mathematics, Rice University, Houston, TX, USA

@@

\section*{Abstract}
% Please keep the abstract between 250 and 300 words
Ambitious projects aim to record the activity of ever larger and denser neuronal populations \emph{in vivo}.  Correlations in neural activity measured in such recordings can reveal important aspects of  neural circuit organization.  However, estimating and interpreting large correlation matrices is statistically challenging.  Estimation can be improved by regularization, \ie by imposing a structure on the estimate.  The amount of improvement depends on how closely the assumed structure represents dependencies in the data. Therefore, the selection of the most efficient correlation matrix estimator for a given neural circuit must be determined empirically.  Importantly, the identity and structure of the most efficient estimator informs about the types of dominant dependencies governing the system.

We sought statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix should be best modeled as the combination of a sparse graph of pairwise partial correlations representing local interactions and a low-rank component representing common fluctuations and external inputs.  Indeed, in cross-validation tests, the covariance matrix estimator with this structure consistently outperformed other regularized estimators. The sparse component of the estimate defined a graph of interactions. These interactions reflected the physical distances and orientation tuning properties of cells: The density of positive \sq{excitatory} interactions decreased rapidly with geometric distances and with differences in orientation preference whereas negative \sq{inhibitory} interactions were less selective.  Because of its superior performance, this \sq{sparse + latent} estimator likely provides a more physiologically relevant representation of the functional connectivity in dense recordings than the sample correlation matrix.

@@

Although the two functions are similar in form, they differ conceptually: The log-likelihood $L(\Sigma|C_{\sf sample})$ is a function of the parameter $\Sigma$ given the sample covariance matrix $C_{\sf sample}$ whereas the loss function $\loss{C,\Sigma}$ expresses the deviation of an arbitrary estimate $C$ from the truth $\Sigma$.

@@

In the example site (Fig.~\ref{fig:4}), the sparse component had 92.8\% sparsity (or conversely, 7.2\% connectivity: $connectivity=1-sparsity$) with average node degree of 20.9 (\figref{4}{G}). The average node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $degree = connectivity\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 72, denoting 72 inferred latent units. The number of latent units increased with population size (\figref{5}{A}) but the average number of connections per cell, \ie \emph{node degree} was highly variable (\figref{5}{B}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and warrants further investigation.

@@

We also examined the relationship of the correlation structure to the differences in orientation preference and to the physical distances between the cell pairs (Fig.\;\ref{fig:6}).  The partial correlations from the $C_{\sf sparse+latent}$ estimates fell more rapidly with the difference in preferred orientation than the sample correlations (\figref{6}{A}). They also decreased more rapidly with physical distance, both laterally (\figref{6}{B}) and in depth (\figref{6}{B}). The faster dropoff may be a trivial result of regularization: $L_1$-shrinkage of the inverse covariance reduces small partial correlations by a greater fraction than large partial correlations.

The positive and negative interactions in the $C_{\sf sparse+latent}$ estimator were organized differently: Consistent with known differences in the spatial distribution synaptic connections formed by excitatory and inhibitory cells \cite{Adesnik:2010,Fino:2011,Perin:2011}, the density of the positive interactions decreased rapidly with differences in the preferred orientation (\figref{6}{D}), lateral displacements (\figref{6}{E}), and displacement in depth (\figref{6}{F}) whereas the negative interactions were much less selective to orientation (\figref{6}{D}), did not decrease in density with lateral distance on the scale of 150 $\mu m$ (\figref{6}{E}), but did decrease with changes in depth, indicating primarily within-layer connectivity (\figref{6}{F}).

@@

Estimator $C_{\sf diag}$ (Eq.~\ref{eq:c-diag})  used two hyperparameters: the covariance shrinkage intensity $\lambda \in [0,1]$ and variance shrinkage intensity $\alpha \in [0,1]$.  The variances (the diagonal of $C_{\sf sample}$) were shrunk toward (linearly mixed with) their mean value:
\begin{equation}
D = (1-\alpha)C_{\sf sample}\circ I + \alpha \frac 1 p \Tr(C_{\sf sample}) I
\end{equation}
Then the diagonal matrix $D$ was used as the target of covariance shrinkage (Eq.~\ref{eq:c-diag}) to produce the final regularized estimate.  The {\tt corpcor} package in the programming language R also implements this estimator \cite{Schafer:2010}, although its analytical optimization of the shrinkage intensities is based on the mean squared error whereas we optimized them with respect to the normal loss function (Eq.~\ref{eq:loss}).

@@

\subsubsection*{Cross-validation with condition-specific variances}
This study required an unbiased estimate of the discrepancy between various correlation matrix estimates and the true value of the correlation matrix. This was accomplished by cross-validation whereby regularized covariance matrix estimates computed from training samples were compared to unbiased covariance matrix estimates computed from independent testing samples. Loss functions were defined with respect to the covariance matrix rather than the correlation matrix because no unbiased estimate of the correlation matrix exists for fininte sample sizes in the general case \cite{Fisher:1921}. In particular, we showed that with the normal loss function (Eq.~\ref{eq:loss}), the validation loss $\loss{C,C_{\sf sample}^\prime}$ is an unbiased estimate of true loss when $C_{\sf sample}^\prime$ is an unbiased estimate. However, this approach only worked under the assumption of uniform variances across all conditions included in the training and testing samples.  Here, we extend the validation loss to the estimation and evaluation of a common correlation matrix across multiple conditions with different variances:

Let $\{x(t):t\in T_c\}$ be a set of measurements collected in condition $c$ from the training sample and let $\{x(t):t \in T_c^\prime\}$ be a set of measurements in condition $c$ from the testing sample, where $T_c$ and $T_c^\prime$ are sets of time samples.

@@

Then the sample correlation matrix in condition $c$ is 
\begin{equation}
	R_{c,\sf sample} = V_{c,\sf sample}^{-\frac 1 2}C_{c,\sf sample}V_{c,\sf sample}^{-\frac 1 2}
\end{equation}
where $V_{c,\sf sample} = C_{c,\sf sample} \circ I$ is the diagonal matrix containing the sample variances.

Then the common correlation matrix $R_{\sf sample}$ is estimated by averaging the condition-specific correlations:
\begin{equation}
	R_{\sf sample} = \frac 1 n\sum\limits_c n_c R_{c,\sf sample}
	= \frac 1 n \sum\limits_c \sum\limits_{t \in T_c} z(t)z(t)^\T
\end{equation}
where $n=\sum\limits_c n_c$, $z(t) = V_{c,\sf sample}^{-\frac 1 2} \left(x(t) - \bar x_c\right)$ is the z-score signal: signal $x(t)$ normalized by the the sample standard deviation specific to the condition. Then $R_{\sf sample}$ is simply the covariance matrix of $z(t)$ across all conditions in the training sample.

With average variances across all conditions $V_{\sf sample} = \frac 1 n \sum\limits_c n_c V_{c,\sf sample}$, the average sample covariance matrix across all conditions is reconstituted from the common sample correlation matrix:
\begin{equation}
	C_{\sf sample} = V_{\sf sample}^{\frac 1 2} R_{\sf sample} V_{\sf sample}^{\frac 1 2}
\end{equation}

The covariance matrix estimators $C_{\sf diag}$, $C_{\sf factor}$, $C_{\sf sparse}$ or $C_{\sf sparse+latent}$ convert $C_{\sf sample}$ into its regularized counterpart denoted here as $C_{\sf reg}$. Then the regularized correlation matrix is
\begin{equation}
	R_{\sf reg} = V_{\sf sample}^{-\frac 1 2}C_{\sf reg}V_{\sf sample}^{-\frac 1 2}
\end{equation}

$R_{\sf reg}$ is not a true correlation matrix because its diagonal values may not equal 1.

The condition-specific variances were regularized toward their mean values across all conditions by linear shrinkage with shrinkage intensity $\delta$:
\begin{equation}
	V_{c,\sf reg} = \delta V_{\sf sample} + (1-\delta) V_{c,\sf sample}
\end{equation}

Then the regularized condition-specific covariance matrices were reconstituted as
\begin{equation}
	C_{c,\sf reg} = V_{c,\sf reg}^{\frac 1 2} R_{\sf reg} V_{c,\sf reg}^{\frac 1 2} 
\end{equation}

The overall loss function is obtained by averaging the loss functions across all the conditions.
With the normal loss function (Eq.~\ref{eq:loss}), this becomes 
\begin{equation}\label{eq:full-loss}
\begin{split}
	&\frac 1 p\sum\limits_c \frac{n_c^\prime}{n^\prime} \loss{C_{c,\sf reg}, C_{c,\sf sample}^\prime}
\\
=& \frac 1 p \sum\limits_c \frac{n_c^\prime}{n^\prime} \left[ \ln\det(C_{c,\sf reg}) + \Tr(C_{c,\sf reg}^{-1}C_{c,\sf sample}^\prime)\right] 
\\
=& \frac 1 p \sum\limits_c \frac{n_c^\prime}{n^\prime} \left[ \ln\det\left(V_{c,\sf reg}^{-\frac 1 2} R_{\sf reg}^{-1} V_{c,\sf reg}^{-\frac 1 2} \right) 
+ \Tr\left(V_{c,\sf reg}^{\frac 1 2} R_{\sf reg} V_{c,\sf reg}^{\frac 1 2} C_{c,\sf sample}^\prime\right)\right] 
\\
=& \frac 1 p \sum\limits_c \frac{n_c^\prime}{n^\prime} \ln \det(V_{c,\sf reg}) 
+ \frac 1 p \ln \det (R_{\sf reg}) + \frac 1 p \Tr\left(R_{\sf reg}^{-1} \sum\limits_c \frac{n_c^\prime}{n^\prime} V_{c,\sf reg}^{-\frac 1 2}C_{c,\sf sample}^\prime V_{c,\sf reg}^{-\frac 1 2}\right)
\\
=& \frac 1 p \sum\limits_c \frac{n_c^\prime}{n^\prime} \Tr(\ln V_{c,\sf reg}) 
+ \frac 1 p \left[\ln \det (\hat R) + \Tr\left(R_{\sf reg}^{-1} R_{\sf sample}^\prime\right)\right] 
\\
=& \frac 1 p \sum\limits_c \frac{n_c^\prime}{n^\prime} \Tr(\ln V_{c,\sf reg}) 
+ \loss{R_{\sf reg}, R_{\sf sample}^\prime}
\end{split}
\end{equation}
Here $n^\prime = \sum\limits_c n_c^\prime$ and
\begin{equation}
R_{\sf sample}^\prime 
= \frac 1 {n^\prime} \sum\limits_c n_c^\prime V_{c,\sf reg}^{-\frac 1 2}C_{c,\sf sample}^\prime V_{c,\sf reg}^{-\frac 1 2}
= \frac 1 {n^\prime} \sum\limits_c \sum\limits_{t \in T_c^\prime}z_{\sf reg}(t) z_{\sf reg}(t)^\T
\end{equation}
where $z_{\sf reg}(t)=V_{c,\sf reg}^{-\frac 1 2}(x-\bar x)$ is the `z-score' trace computed by normalizing the signals from the testing sample by the regularized variance estimates from the training sample. $R_{\sf sample}^\prime$ is not a true correlation matrix because its diagonal elements may not equal 1.

The loss function in Eq.~\ref{eq:full-loss} was used in the main results of this study when the correlation matrix was estimated across multiple stimulus conditions. The variance regularization parameter $\delta$ was optimized in the nested cross validation loop together with the covariance estimation hyperparameters.

>>>>

\begin{document}
% Title must be 150 characters or less
\begin{flushleft}
{\Large
Improved estimation and interpretation of correlations in neural circuits
}
% Insert Author names, affiliations and corresponding author email.
\\
Dimitri Yatsenko$^{1}$, 
Kre\v{s}imir Josi\'{c}$^{2}$,
Alexander S.~Ecker$^{1,3,4}$,
Emmanouil Froudarakis$^{1}$,
R.~James Cotton$^{1}$,
Andreas S.~Tolias$^{1,5,\ast}$
\\
\bf{1} Department of Neuroscience, Baylor College of Medicine, Houston, TX, USA
\\
\bf{2} Department of Mathematics and Department of Biology and Biochemistry, University of Houston, Houston, TX, USA
\\
\bf{3}  Werner Reichardt Center for Integrative Neuroscience and Institute for Theoretical Physics, University of T\"ubingen, Germany
\\
\bf{4} Bernstein Center for Computational Neuroscience, T\"ubingen, Germany
\\
\bf{5} Department of Computational and Applied Mathematics, Rice University, Houston, TX, USA

@@

\section*{Abstract}
% Please keep the abstract between 250 and 300 words
Ambitious projects aim to record the activity of ever larger and denser neuronal populations \emph{in vivo}.  Correlations in neural activity measured in such recordings can reveal important aspects of  neural circuit organization.  However, estimating and interpreting large correlation matrices is statistically challenging.  Estimation can be improved by regularization, \ie by imposing a structure on the estimate.  The amount of improvement depends on how closely the assumed structure represents dependencies in the data. Therefore, the selection of the most efficient correlation matrix estimator for a given neural circuit must be determined empirically.  Importantly, the identity and structure of the most efficient estimator informs about the types of dominant dependencies governing the system.
We sought statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix should be best modeled as the combination of a sparse graph of pairwise partial correlations representing local interactions and a low-rank component representing common fluctuations and external inputs.  Indeed, in cross-validation tests, the covariance matrix estimator with this structure consistently outperformed other regularized estimators. The sparse component of the estimate defined a graph of interactions. These interactions reflected the physical distances and orientation tuning properties of cells: The density of positive \sq{excitatory} interactions decreased rapidly with geometric distances and with differences in orientation preference whereas negative \sq{inhibitory} interactions were less selective.  Because of its superior performance, this \sq{sparse + latent} estimator likely provides a more physiologically relevant representation of the functional connectivity in dense recordings than the sample correlation matrix.

@@

Although the two functions are similar in form, they differ conceptually: The log-likelihood $L(\Sigma|C_{\sf sample})$ is a function of the parameter $\Sigma$ given the sample covariance matrix $C_{\sf sample}$ whereas the loss function $\loss{C,\Sigma}$ expresses the deviation of an arbitrary estimate $C$ from the truth $\Sigma$. Minimum expected loss may be attaned at a point far removed from the point of maximum likelihood point given the same sample even when the normal distribution can be assumed.

@@

In the example site (Fig.~\ref{fig:4}), the sparse component had 92.8\% sparsity (or conversely, 7.2\% connectivity: $\mbox{connectivity}=1-\mbox{sparsity}$) with average node degree of 20.9 (\figref{4}{G}). The average node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $\mbox{degree} = \mbox{connectivity}\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 72, denoting 72 inferred latent units. The number of latent units increased with population size (\figref{5}{A}) but the average number of connections per cell, \ie \emph{node degree} was highly variable (\figref{5}{B}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and warrants further investigation.

@@

We examined the relationship of the structure of the $C_{\sf sparse+latent}$ estimates to the differences in orientation preference and to the physical distances separating pairs of cells (Fig.\;\ref{fig:6}).  Five sites with highest connectivitites were included in the analysis. Partial correlations were computed using Eq.~\ref{eq:partial} based on the regularized estimate, including both the sparse and the latent component. Connectivity was computed as the fraction of pairs of cells in each categority connected by non-zero elements (interactions) in the sparse component of the estimate, distinguishing between the positive and negative connectivities.

First, we compared how correlations and connectivity depended on the difference in preferred orientations of pairs of significantly ($\alpha=0.05$) tuned cells. The partial correlations decreased more rapidly with differences in tuning than sample correlations ($p<10^{-9}$ in each of the five sites, two-sample $t$-test on the linear regression coefficients). Positive connectivity decreased rapidly with difference in the preferred orientations ($p<0.005$ in each of the five sites, $t$-test on the logistic regression parameter) whereas negative connectivity did not increase (\figref{6}{D}): The dependence of connectivity on difference in orientation was significantly higher for positive than for negative interactions ($p<0.04$ in each of the five sites, two-sample $t$-test on the logistic regression parameters).

Second, we compared how correlations and connectivity depended on the physical distance separating pairs of cells. We distinguished between lateral displacement, \ie the distance in the \emph{xy} plane between cell pairs located at similar depths $\Delta z < 30 \mu m$, and depth displacement, \ie the difference in depth with lateral displacement under $30\,\mu m$. Again, the partial correlations decreased more rapidly with both lateral and depth displacements ($p<10^{-6}$ in each of the five sites, both lateral and depth, two-sample $t$-test on the linear regression coefficients). 
Lateral positive connectivity decreased rapidly with distance ($p<10^{-6}$ in each of the five sites, $t$-test on the logistic regression parameter) (\figref{6}{E}), so that cells separated laterally by less than 25 $\mu m$ were 3.2 times more likely to be connected than cells separated laterally by more than 150 $\mu m$. The negative connectivity did not significantly change with lateral displacement (\figref{6}{E}): The dependence of connectivity on lateral displacement was significantly higher for positive than for negative interactions ($p<0.05$ in each of the five sites, two-sample $t$-test on the logistic regression parameters). Positive connectivity decreased significantly with increased depth displacement of pairs of cells ($p<0.05$ in each of the five sites, $t$-test on the logistic regression parameter).  Negative connectivity on depth displacement did not reach significance in most sites using the same regression test nor could this dependence be shown to be sufficiently different from the positive connectivity.

@@

Estimator $C_{\sf diag}$ (Eq.~\ref{eq:c-diag})  used two hyperparameters: the covariance shrinkage intensity $\lambda \in [0,1]$ and variance shrinkage intensity $\alpha \in [0,1]$.  The variances (the diagonal of $C_{\sf sample}$) were shrunk toward (linearly mixed with) their mean value:
\begin{equation}
D = (1-\alpha)C_{\sf sample}\circ I + \alpha \frac 1 p \Tr(C_{\sf sample}) I
\end{equation}
Then the diagonal matrix $D$ was used as the target of covariance shrinkage (Eq.~\ref{eq:c-diag}) to produce the final regularized estimate.

@@

\paragraph{Cross-validation with condition-specific variances}
This study required an unbiased estimate of the discrepancy between various correlation matrix estimates and the true value of the correlation matrix. This was accomplished by cross-validation whereby regularized covariance matrix estimates computed from training samples were compared to unbiased covariance matrix estimates computed from independent testing samples. Loss functions were defined with respect to the covariance matrix rather than the correlation matrix because no unbiased estimate of the correlation matrix exists for fininte sample sizes in the general case \cite{Fisher:1921}. In particular, we showed that with the normal loss function (Eq.~\ref{eq:loss}), the validation loss $\loss{C,C_{\sf sample}^\prime}$ is an unbiased estimate of true loss when $C_{\sf sample}^\prime$ is an unbiased estimate. However, this approach only worked under the assumption of uniform variances across all conditions included in the training and testing samples.  Here, we extend the validation loss to the estimation and evaluation of a common correlation matrix across multiple conditions with different variances:

Let $\{x(t):t\in T_c\}$ be a set of measurements collected in condition $c$ from the training sample and let $\{x(t):t \in T_c^\prime\}$ be a set of measurements in condition $c$ from the testing sample, where $T_c$ and $T_c^\prime$ are time bin indices for the training and testing samples, respectively. We marked all quantities estimated from the testing sample with a prime mark: \emph{e.g.}\,$C_{c,\sf sample}^\prime$, $n_c^\prime$.

@@

To apply regularization to the correlation matrix, we first reconstruct the common covariance matrix.
The common correlation matrix $R_{\sf sample}$ is estimated by averaging the condition-specific correlations:
\begin{equation}
	R_{\sf sample}  
	= \frac 1 n \sum\limits_c n_c \left(V_{c,\sf sample}^{-\frac 1 2}C_{c,\sf sample}V_{c,\sf sample}^{-\frac 1 2}\right)
        = \frac 1 n \sum\limits_c \sum\limits_{t \in T_c} z(t)z(t)^\T,
\end{equation}
where $n=\sum\limits_c n_c$ and $V_{c,\sf sample} = C_{c,\sf sample} \circ I$ is the diagonal matrix containing the sample variances. Then $R_{\sf sample}$ is simply the covariance matrix of the $z$-score signal 
$z(t) = V_{c,\sf sample}^{-\frac 1 2} \left(x(t) - \bar x_c\right)$ of the training sample.

For consistency with prior work, we applied regularization to covariance matrices rather than to correlation matrices. The common covariance matrix was estimated by scaling $R_{\sf sample}$ by the average variances across conditions $V_{\sf sample} = \frac 1 n \sum\limits_c n_c V_{c,\sf sample}$:
\begin{equation}
	C_{\sf sample} = V_{\sf sample}^{\frac 1 2} R_{\sf sample} V_{\sf sample}^{\frac 1 2}
\end{equation}
Note that $C_{\sf sample}$ differs from the sample covariance matrix computed without conditioning the variances on $c$ and this computation helps avoid any biases that would be introduced by ignoring changes in variance.

The covariance matrix estimators $C_{\sf diag}$, $C_{\sf factor}$, $C_{\sf sparse}$ or $C_{\sf sparse+latent}$ convert $C_{\sf sample}$ into its regularized counterpart denoted here as $C_{\sf reg}$.

To evaluate the estimators, we regularized the conditioned variances by linear shrinkrage toward their mean value across all conditions. This was accomplished by scaling $C_{\sf reg}$ by the diagonal variance adjustment matrix 
\begin{equation}
	Q_c = \delta I + (1-\delta)V_{\sf sample}^{-1} V_{c,\sf sample}
\end{equation}
where $\delta \in [0,1]$ is the variance regularization hyperparameter, which was optimized in the inner loop of cross-validation along with the regularization hyperparameters for $C_{\sf reg}$.

Then the regularized condition-specific covariance matrices became
\begin{equation}
	C_{c,\sf reg} = Q_c^{\frac 1 2} C_{\sf reg} Q_c^{\frac 1 2} 
\end{equation}

The overall loss function is obtained by averaging the loss functions across all the conditions:
\begin{equation}\label{eq:full-loss}
	\frac 1 {\sum\limits_c n_c^\prime}\sum\limits_c n_c^\prime \loss{C_{c,\sf reg}, C_{c,\sf sample}^\prime} 
\end{equation}

With the normal loss $\loss{\cdot,\cdot}$ (Eq.~\ref{eq:loss}) and the unbiased testing covariance matrix $C_{c,\sf sample}$, the loss function in Eq.~\ref{eq:full-loss} is an unbiased estimate of the true loss. Hence, it was used for evaluations reported in \figref{3}.
